Formation and evolution of intermediate-mass ratio inspirals.

### 2020-07-07 main.tex
ready for resubmission

the simulations output, in Section \ref{gra} we discuss the implications for GW
astronomy, while Section \ref{res} is devoted to the results summary and
conclusions.

@@

We vary the central IMBH mass in four classes, namely ${\rm Log}(M_\ibh/\Ms) = 2-3-4-5$. This range covers typical values of putative IMBH masses forming in stellar systems of various sizes, from young and open clusters, to globular clusters, and up to low-mass nuclear clusters and dwarf galaxies.
We note that the lowest value taken for $M_\ibh$ will likely involve mergers with mass ratio $> 0.1$, thus they fall outside the range of IMRIs. Nonetheless, exploring the low end of IMBH mass function will help us in better understanding the perspectives of IMBH-BH mergers from the point of view of both low- and high-frequency GW detectors, as discussed in Section XX.

The properties of the cluster in which the IMBH is embedded, which define the cluster potential, are varied depending on the model. The cluster density profile is assumed to be either a Dehnen sphere with inner slope $\gamma_\gc = 0.5$ (model S0, S3, S4, S5, S6) or $1.0$ (model S1), or a Plummer sphere (model S2). In both cases, the cluster half-light radius is assumed to be $R_{\rm eff} = 3.4$ pc, namely the mean value of Milky Way globular clusters \citep{harris14}. The mass of the cluster is calculated via the scaling provided by \cite{AS16}, connecting the host cluster mass $M_\gc$ with the total ``dark'' mass, inhabiting the cluster's centre, comprised of either an IMBH or a sizable population of stellar BHs
\begin{equation}
\Log \left(\frac{M_\gc}{\Ms}\right) = \alpha \Log \left(\frac{M_\ibh}{\Ms}\right) + \beta.
\label{mcmibh}
\end{equation}
with $\alpha = 0.999 \pm 0.001$ and $\beta = 2.23 \pm 0.009$. The cluster typical radius is thus calculated from the assumed $R_{\rm eff}$ and the adopted mass profile. Upon these assumptions an IMBH with mass $M_\ibh = 10^2(10^5)\Ms$ is associated with a cluster mass of $ M_\gc = 1.7\times 10^4(1.7\times 10^7)\Ms$, thus our models ideally span the mass range of young clusters, globular clusters, and dwarf galaxies nuclei. We assume that the IMBH is orbited by two stellar BHs, since the IMBH is expected to be the most massive object in the cluster and the dominant element in determining the dynamics. The orbital eccentricity $e_{1,2}$ is extracted randomly according to a thermal distribution $P(e){\rm d} e = 2e{\rm d}e$, while the semimajor axis $a_{1,2}$ is selected according to the overall cluster density profile with the exception of model S6, in which we extract $a_{1,2} = 0.1-10^4$ AU assuming a distribution flat in logarithmic values.  
In models S0 to S5, the maximum semimajor axis allowed is set equal to the radius within which the host cluster contains two stellar BHs with average mass $30\Ms$.

@@

{\bf 
All simulations are performed taking advantage of \ARGdf \citep{ASCD19}, a modified version of the \ARCHAIN code that implements post-Newtonian (PN) dynamics and {\it algorithmic regularization} to handle close encounters and strong collisions \citep{mikkola99,mikkola08}. For our purposes, we include in our treatment only 1, 2, and 2.5 order PN terms. Additionally, \ARGdf allows the user to take into account in particles' equations of motion the gravitational field generated by the host stellary system and a dynamical friction term. Simulations are halted either if one of the two BH merge with the IMBH, if one of the BHs is ejected away if the simulated time exceeds $t = 1$ Gyr, or if the runtime exceeded 1 hr. The $1$ Gyr time limit serves to cope with the fact that IMBH in the range of masses $>10^3\Ms$ are expected to build-up via stellar accretion and mergers with compact remnants over timescales that can range considerably $1-10$ Gyr depending on the formation pathway \citep[see for instance][]{giersz15}. A summary of all models considered is provided in Table \ref{tabt1}.
}

@@

Clearly, while the categorization provided suggests three well separate classes, it must be noted that models falling in case c) (mergers) might have undergone a chaotic triple phase, or secular effects that triggered the IMBH-BH merger. In fact, figure \ref{fig:example} shows a typical example of one of the simulations performed in which the merger takes place within the simulated time. In the example, an IMBH with mass $M_\ibh = 10^2\Ms$ forms a tight binary with a stellar BH with mass $M_\bh = 15\Ms$, the binary is subjected to perturbation of the outer BH with mass $M_\bh = 5\Ms$ that causes a continuous oscillation of the eccentricity up to the point -- around 300 Myr from the beginning of the simulation -- at which the eccentricity peaks at values $\sim 0.999998$ and GW start dominating the IMBH-BH binary evolution which culminate with a merger shortly after.  
}
\begin{figure}
\includegraphics[width=\columnwidth]{example}
    \caption{Time evolution of the semimajor axis (straight blue line) and eccentricity (dotted red line) for one of the simulations performed in SET0.}
	\label{fig:example}
\end{figure}

The amount of both ``direct'' and ``delayed'' merger depends, as outlined in Table \ref{tab:t2}, on the IMBH mass and the model investigated. Figure \ref{fig:gwprob} shows the percentage of total mergers (direct + delayed, hereafter $f_{\rm GW}$) as a function of the IMBH mass for different models. We see that the scatter among different models and for a fixed IMBH mass value is considerable, spanning from $f_{\rm GW} = 10\%$ (SET 2) to $50 \%$ (SET 6) for $M_\ibh = 10^3\Ms$. 
\begin{figure}
\includegraphics[width=\columnwidth]{mergerprob}
    \caption{IMRIs merger probability as a function of the IMBH mass for all models explored.}
	\label{fig:gwprob}
\end{figure}

@@

Another important property that can be inferred from our models is the mass distribution ${\rm d}N/{\rm d}M_{\rm bh}$ of IMRI lighter companion, which is shown in Figure \ref{fig:f8} for models SET0-5 (different cluster density profiles and BH mass spectrum) and Figure \ref{fig:f8} for models SET0, 6 (different stellar metallicity). Comparing SET0, 1, and 3 -- which differ only in the cluster density profile but adopt the same BH mass spectrum \citep{spera17} -- it is apparent that the ${\rm d}N/{\rm d}M_{\rm bh}$ not depend on the environment, but rather on the BH mass spectrum adopted. In all three cases, the mass distribution shows a clearly bimodal distribution peaked at $\sim 7\Ms$ and $34\Ms$. The picture changes significantly if another BH mass spectrum is adopted. In the case of a power-law mass spectrum in the range $M_\bh = 3-30\Ms$ we find that ${\rm d}N/{\rm d}M_{\rm bh}$ declines sharply starting from $3\Ms$ and truncates at $> 25 \Ms$, whereas for a flat mass spectrum in the same range the trend turns out to be opposite, with the ${\rm d}N/{\rm d}M_{\rm bh}$ parameter increasing toward $25 \Ms$ and abruptly dropping beyond this value. Assuming smaller values for the initial semimajor axis (SET5) does not affect significantly the merging BH mass distribution, which in fact resembles that obtained for SET0-3. We explored whether the IMBH mass plays a role in shaping the merging BH mass distribution but, on average, we find that the overall distribution remains the same across all the range of $M_\ibh$ explored. As shown in Figure \ref{fig:mbhdist}, increasing the stellar metallicity to solar values (SET6) leads to IMRIs with a secondary component mass that remains below $25\Ms$.

These results have two main implications. On the one hand, the IMRI secondary mass tells us something about the mass spectrum of BHs lurking around an IMBH. On the other hand, it provides clues about the IMBH formation process. In fact, if the IMBH formed during cluster's early life, we can expect that the BH mass spectrum isn't altered substantially by dynamics. However, if the IMBH develops after dynamical encounters shaped the BH reservoir -- for instance by causing the ejection of the most massive BHs -- we would expect that the IMRI secondary mass would preferentially attains lower values.

@@

\subsection{IMBH survival in dense star clusters}
{\bf
Promptly after a merger event, the anysotropic emission of GWs can impart a recoil kick to the merger product, depending on the mass and spin of the two merging objects \citep{campanelli07,gonzalez07,lousto08,lousto12}. Depending on the kick amplitude, the recoil can kick out the IMBH from the cluster \citep[e.g.][]{bockelmann08}, leading it to wander forever outside the cluster. In fact, for low-velocity dispersion ($5$ km/s) star clusters the probability to retain a BH with mass $m_\bh> 100 \Ms$ after 1-2 consecutive mergers is rather low, $1-5\%$ \citep{arca20}. To place constraints on the IMBH retention probability, we create a catalogue of IMRI mergers proceeding on a step-by-step basis as follows: 
\begin{enumerate}
\item we divide the IMBH mass range -- $M_\ibh = 10^2-5\times 10^5\Ms$ -- in 15 values evenly distributed in logarithmic values;
\item for each IMBH mass, we create a sample of 100 stellar BHs with masses evenly drawn in the $M_\bh = 3.5-80 \Ms$ range;
\item we associate to the IMBH and BH a spin randomly drawn between 0-1;
\item we use \cite{jimenez17} numerical relativity fitting formulae to calculate the IMRI merger remnant mass $M_r$, spin $S_r$, and recoil velocity $v_r$, following the procedure depicted in \cite{arca20}.
\end{enumerate}
For each $M_\ibh -m_\bh$ pair, we repeat steps 3 and 4 100 times and, for each of them, we compare the recoil velocity with the escape velocity from the cluster centre, which we set to $\sigma = 5$ or $50$ km/s.

The result of this procedure is described in Figure \ref{fig:f11}, which shows the percentage of models for which $v_r < \sigma$. In the case of 
a ''normal'' cluster ($\sigma = 5$ km/s), we find that an IMBH with mass $M_\ibh < 200 \Ms$ gets ejected whenever the merging companion has a mass $m_\bh \geq 10 \Ms$. Even for heavier IMBHs, e.g. $M_\ibh = 1000\Ms$, the retention probability rapidly drops below $50\%$  if the companion mass exceeds $M_\bh > 20\Ms$. Only for quite heavy IMBHs ($M_\ibh \sim 10^4\Ms$) the retention probability attains values $>80\%$ regardless the secondary BH mass. In the case of a denser cluster ($\sigma=50$ km/s), we find that a moderate IMBH ($M_\ibh \sim 200-300 \Ms$) have a $50\%$ retention probability if $m_\bh < 30\Ms$, while for $M_\ibh = 1000\Ms$ the BH mass threshold to ensure a $50\%$ retaining probability increases to $m_\bh = 70\Ms$.

Therefore, localizing an IMBH in the centre of a star cluster and measuring its mass would enable us to get insights on its evolution. For instance, the presence of an IMBH with mass $>10^3\Ms$, even in dense clusters, would suggest that the IMBH growth process might be dominated  by matter accretion from disrupted star rather than by mergers with BHs heavier than $30-50\Ms$.
}

@@

\subsection{IMBHs spin evolution}
{\bf 
Since an IMBH-BH merger cause an inevitable variation of the IMBH spin, this quantity might be useful to get insights on the IMBH evolution.
To support this idea, we create a sample of IMBH-BH mergers as follows. We select an IMBH with mass of either $M_\ibh = 10^2 - 10^3 - 10^4 - 10^5$ and spin $S_\ibh = 0.1 - 0.5 - 0.9$ and we merge it with a stellar BH with mass in the range $3 - 60 \Ms$ drawn randomly and spin drawn accordingly to a Gaussian distribution with mean $S_\bh = 0.7$ -- i.e. the average value inferred from LIGO O1+O2 observations -- and dispersion 0.1.

@@

We see that as long as the mass ratio remains below $10^{-2}$ -- which implies an $M_\ibh > 10^4 \Ms$ -- the IMBH spin is not affected by the merger. However, for IMRIs with a mass ratio > 0.01, the remnant IMBH spin can change significantly. For mass ratios in the range $q = 0.01-0.1$ the spin variation is $\sim \pm 0.1$ with respect to the initial value, whereas at larger mass ratio values we see that the spin tend to move toward values close to the average value assumed for the BH spin distribution, being the remnant spin $S_r \sim 0.8\pm 0.2$.
The peculiar dependence between the remnant spin and the IMRI mass ratio has important consequences for the IMBH evolution. One the one hand, if the IMBH formed from the collapse of a gaseous cloud or a very massive star born out of stellar collisions and reach masses $\gtrsim 10^4\Ms$, measuring its spin in a IMRI merger would provide us with information on the processes that regulated the IMBH very formation. On the other hand,
if the IMBH formed via multiple mergers in a stellar environment rich in BHs, the remnant spin can provide clues on the spin distribution of the underlying population of stellar BHs.

To further investigate this point, we explore the effects of a series of subsequent mergers on the IMBH final spin in the case of 
an IMBH mass value  that ensure a retention probability of at least $50\%$ in an IMRI merger event involving a BH smaller than $40\Ms$.
From Figure \ref{fig:survival} and Figure\ref{fig:f11a} we thus select $M_\ibh = 10^3\Ms$, which satisfies the requirement above, and set its spin to either $S_\ibh = 0.1-0.5-0.9$. For each spin value, we model ten consecutive mergers with stellar BHs with masses $3-60\Ms$ and spin drawn from a Gaussian distribution with average $S_\bh = 0.5$ and dispersion 0.1. This sequence represents an {\it IMBH merger tree}. For each value of the spin we create 100 merger trees and calculate the IMBH spin evolution as shown in Figure \ref{fig:f11b}.
We note that all models tend to converge toward the mean value of the adopted spin distribution for stellar BHs, thus suggesting that it could be possible to link the IMBH spin with the overall spin distribution of stellar BHs inhabiting the star cluster. Infact, complementing a measure of the IMBH spin with the spin distribution inferred from stellar BH mergers observed as GW sources can help in unravelling the IMBH evolution, whether it was driven by mergers with compact objects or by stellar feeding.
}

@@

\subsection{Gravitational wave signal}
{\bf
Our simulations encompass a wide range of IMRI models, in fact touching the layer of ``ordinary'' BH binaries (mass ratio $q>0.1$) and 
scratching the limit of extreme-mass ratio inspirals ($q < 3 \times10^{-5}$). This setup implies that the GW emission connected with our modelled mergers can populate a wide range of frequencies, from milli- to deca-Hz. Figure \ref{fig:f13} show the characteristic amplitude as a function of the frequency for a sample of IMRI mergers with different IMBH mass, assuming they are located at redshift $z=0.1$ ($\sim 460$ Mpc) and an observation time of 4 yr. We compare the modelled signal with the sensitivity curve of several detectors, namely LISA, ALIA, DO, DECIGO, LIGO, and the Einstein Telescope (ET). We see that mergers involving IMBHs with masses $< 10^5\Ms$ are promising multiband sources that can be seen in one detector during the inspiral phase and in another during the merger. The number of regions in which a detector is sensitive traversed by an IMRI widens at increasing the IMBH mass, in fact in principle an IMBH with mass $<10^3\Ms$ can be seen across the whole milli-decaHz frequency range.

@@

\subsection{Eccentricity of IMRIs}
{\bf
One important parameter that could be inferred from GW observations is the eccentricity of the source. To better understand whether our IMRI models are expected to retain a significant eccentricity while transiting across different observational bands, we show in Figure \ref{fig:f10} the average value of the eccentricity calculated when IMRIs cross the milli-Hz, deci-Hz, and Hz bands for all simulations set. It is apparent that even when emitting low-frequency GWs, IMRIs have already significantly circularized, although apparently lower-mass IMRIs tend to have slightly larger eccentricities. The fact that our IMRIs are substantially circular when becoming observable to GW detectors has implications on their detectability, as we show in the next section.
}

@@

{\bf In fact, in the last decade a number of works pointed out the potential presence of IMBHs in Galactic globular clusters, although in most cases the results were inconclusive. Concerning clusters orbiting closer than 8 kpc to us candidates include: 47Tuc \citep{kiziltan17}, NGC6266 \citep{abbate19}, NGC6128 and NGC288 \citep{sollima16}, NGC6388 and NGC2808 \citep{miocchi07,lanzoni07}.
For instance, the inferred value for the 47Tuc IMBH mass goes from $M_\ibh > 2,000\Ms$ \citep{kiziltan17} to substantially lower values $\ll 1,000\Ms$ \citep{abbate19b}, depending on the observational technique adopted. A similar discussion involves other clusters, like NGC6388 \citep{lanzoni13,lutzgendorf13}. Therefore, LISA might play a crucial role in unravelling the presence (or not) of IMBHs in the Galaxy, at least the closest one.

Not only, in fact LISA can also be used to provide clues about the presence of IMBHs in nearby galaxies. For instance, it has been recently suggested that the Large Magellanic Cloud might be harbouring an IMBH with a mass in the range $4\times 10^3-10^4\Ms$ \citep{erkal19} and up to $10^7\Ms$ \citep{boyce17}, and that some of the clusters residing in the cloud can host IMBHs as well with masses in the range $10^3-10^4\Ms$ \citep{gualandris07}. Assuming a distance to the LMC $D_L = 49.97 \pm 1.13$ kpc \citep{pietrzy13}, we find that LISA might deliver observations of {\it Magellanic} IMRIs with an $SNR = 8-40$, with the lower(upper) value corresponding to an IMBH with mass $10^3\Ms$($10^4\Ms$) and a BH companion of $30\Ms$. 
}

>>>>

the simulations output and the implications for IMBH formation and evolution, in Section \ref{gra} we discuss the implications for GW astronomy, while Section \ref{end} is devoted to the results summary and conclusions.

@@

We vary the central IMBH mass in four classes, namely ${\rm Log}(M_\ibh/\Ms) = 2-3-4-5$. This range covers typical values of putative IMBH masses forming in stellar systems of various sizes, from young and open clusters, to globular clusters, and up to low-mass nuclear clusters and dwarf galaxies.
We note that the lowest value taken for $M_\ibh$ will likely involve mergers with mass ratio $> 0.1$, thus they fall outside the range of IMRIs. Nonetheless, exploring the low end of IMBH mass function will help us in better understanding the perspectives of IMBH-BH mergers from the point of view of both low- and high-frequency GW detectors, as discussed in Section \ref{gra}.

The properties of the cluster in which the IMBH is embedded, which define the cluster potential, are varied depending on the model. The cluster density profile is assumed to be either a Dehnen sphere with inner slope $\gamma_\gc = 0.5$ (model S0, S3, S4, S5, S6) or $1.0$ (model S1), or a Plummer sphere (model S2). In both cases, the cluster half-light radius is assumed to be $R_{\rm eff} = 3.4$ pc, namely the mean value of Milky Way globular clusters \citep{harris14}. The mass of the cluster is calculated via the scaling provided by \cite{AS16}, connecting the host cluster mass $M_\gc$ with the total ``dark'' mass, inhabiting the cluster's centre, comprised of either an IMBH or a sizable population of stellar BHs
\begin{equation}
\Log \left(\frac{M_\gc}{\Ms}\right) = \alpha \Log \left(\frac{M_\ibh}{\Ms}\right) + \beta.
\label{mcmibh}
\end{equation}
with $\alpha = 0.999 \pm 0.001$ and $\beta = 2.23 \pm 0.009$. The cluster typical radius is thus calculated from the assumed $R_{\rm eff}$ and the adopted mass profile. Upon these assumptions an IMBH with mass $M_\ibh = 10^2(10^5)\Ms$ is associated with a cluster mass of $ M_\gc = 1.7\times 10^4(1.7\times 10^7)\Ms$, thus our models ideally span the mass range of young clusters, globular clusters, nuclear clusters, and dwarf galaxies nuclei. We assume that the IMBH is orbited by two stellar BHs, since the IMBH is expected to be the most massive object in the cluster and the dominant element in determining the dynamics. 
The choice of limiting the number of BH companions to two is dictated by the numerical evidence that IMBHs tend to form after the reservoir of BH population diminished severely through dynamical interactions \citep{zwart02,giersz15}. Compared to other works focused on a similar topics, our models do not rely on any  preferential configuration or initial hierarchy for the IMBH-BH-BH system, since we expect that the evolution around the IMBH will be mostly driven by the chaotic interactions involving the compact objects surrounding the IMBH \citep[e.g.][]{konstantinidis13}. The orbital eccentricity $e_{1,2}$ is extracted randomly according to a thermal distribution $P(e){\rm d} e = 2e{\rm d}e$, while the semimajor axis $a_{1,2}$ is selected according to the overall cluster density profile with the exception of model S6, in which we extract $a_{1,2} = 0.1-10^4$ AU assuming a distribution flat in logarithmic values. This allows us to explore the role of the BHs spatial distribution in determining the IMRIs formation and merger.

In models S0 to S5, the maximum semimajor axis allowed is set equal to the radius within which the host cluster contains two stellar BHs with average mass $30\Ms$.

@@

{\bf 
All simulations are performed taking advantage of \ARGdf \citep{ASCD19}, a modified version of the \ARCHAIN code that implements post-Newtonian (PN) dynamics and {\it algorithmic regularization} to handle close encounters and strong collisions \citep{mikkola99,mikkola08}. For our purposes, we include in our treatment only 1, 2, and 2.5 order PN terms. Additionally, \ARGdf allows the user to take into account in particles' equations of motion the gravitational field generated by the host stellary system and a dynamical friction term. Simulations are halted either if: one of the two BH merge with the IMBH, one of the BHs is ejected away, the simulated time exceeds $t = 1$ Gyr, or the runtime exceeded 1 hr. The $1$ Gyr time limit serves to cope with the fact that IMBH in the range of masses $>10^3\Ms$ are expected to build-up via stellar accretion and mergers with compact remnants over timescales that can range considerably $1-10$ Gyr depending on the formation pathway \citep[see for instance][]{giersz15}. A summary of all models considered is provided in Table \ref{tabt1}.
}

@@

Clearly, while the categorization provided suggests three well separate classes, it must be noted that models falling in case c) (mergers) might have undergone a chaotic triple phase, or secular effects that triggered the IMBH-BH merger. Figure \ref{fig:example} shows a typical example of one of the simulations performed in which the merger takes place within the simulated time: an IMBH with mass $M_\ibh = 10^2\Ms$ forms a tight binary with a stellar BH with mass $M_\bh = 15\Ms$, the binary is subjected to perturbation of the outer BH with mass $M_\bh = 5\Ms$ that causes a continuous oscillation of the eccentricity up to the point -- around 300 Myr from the beginning of the simulation -- at which the eccentricity peaks at values $\sim 0.999998$, GW emission kicks in and start dominating the IMBH-BH binary evolution eventually culminating in a merger.  
}
\begin{figure}
\includegraphics[width=\columnwidth]{example}
    \caption{Time evolution of the semimajor axis (straight blue line) and eccentricity (dotted red line) for one of the simulations performed in SET0.}
	\label{fig:example}
\end{figure}

{\bf 
As outlined in Table \ref{tab:t2}, the amount of both ``direct'' and ``delayed'' merger depends on the IMBH mass and the model adopted. Figure \ref{fig:gwprob} shows the percentage of total mergers (direct + delayed, hereafter $f_{\rm GW}$) as a function of the IMBH mass for different models. We see that the scatter among different models and for a fixed IMBH mass value is considerable, spanning from $f_{\rm GW} = 10\%$ (SET 2) to $50 \%$ (SET 6) for $M_\ibh = 10^3\Ms$. 
\begin{figure}
\includegraphics[width=\columnwidth]{mergerprob}
    \caption{IMRIs merger probability as a function of the IMBH mass for all models explored.}
	\label{fig:gwprob}
\end{figure}

@@

Another important property that can be inferred from our models is the mass distribution ${\rm d}N/{\rm d}M_{\rm bh}$ of the IMRI secondary, which is shown in Figure \ref{fig:f8} for models SET0-5 (different cluster density profiles and BH mass spectrum) and Figure \ref{fig:mbhdist} for models SET0, 6 (different stellar metallicity). Comparing SET0, 1, and 3 -- which differ only in the cluster density profile but adopt the same BH mass spectrum \citep{spera17} -- it is apparent that the ${\rm d}N/{\rm d}M_{\rm bh}$ not depend on the environment, but rather on the BH mass spectrum adopted. In all three cases, the mass distribution shows a clearly bimodal distribution peaked at $\sim 7\Ms$ and $34\Ms$. The picture changes significantly if another BH mass spectrum is adopted. In the case of a power-law mass spectrum in the range $M_\bh = 3-30\Ms$ we find that ${\rm d}N/{\rm d}M_{\rm bh}$ declines sharply starting from $3\Ms$ and truncates at $> 25 \Ms$, whereas for a flat mass spectrum in the same range the trend turns out to be opposite, with the ${\rm d}N/{\rm d}M_{\rm bh}$ parameter increasing toward $25 \Ms$ and abruptly dropping beyond this value. Assuming smaller values for the initial semimajor axis (SET5) does not affect significantly the merging BH mass distribution, which in fact resembles that obtained for SET0-3. We explored whether the IMBH mass plays a role in shaping the merging BH mass distribution but, on average, we find that the overall distribution remains the same across all the range of $M_\ibh$ explored. As shown in Figure \ref{fig:mbhdist}, increasing the stellar metallicity to solar values (SET6) implies a reduction of the maximum value of the IMRI secondary mass to $25\Ms$.

These results have two main implications. On the one hand, the IMRI secondary mass tells us something about the mass spectrum of BHs lurking around an IMBH. On the other hand, it provides clues about the IMBH formation process. In fact, if the IMBH formed during cluster's early life, we might expect that the mass spectrum of IMRIs secondary maps roughly the stellar BH mass spectrum. If, instead, dynamical interactions have enough time to shape the BHs reservoir, we might expect that the mass spectrum of IMRI secondaries will reflect the outcome of stellar dynamics.

@@

\subsection{IMBH survival in dense star clusters}
{\bf
Promptly after a merger event, the anysotropic emission of GWs can impart a recoil kick to the merger product, depending on the mass and spin of the two merging objects \citep{campanelli07,gonzalez07,lousto08,lousto12}. Depending on the kick amplitude, the recoil can kick out the IMBH from the cluster \citep[e.g.][]{bockelmann08}, leading it to wander forever outside the cluster. 
This can significantly affect the retention probability of low-velocity dispersion ($5$ km/s) star clusters, which is limited to $1-5\%$ for BHs with mass $m_\bh \sim 100 \Ms$ that undergo 1-2 consecutive mergers \citep{arca20}.

Assessing the retention probability for IMBHs is crucial to place constraints on their possible presence in star clusters. To explore this aspect, we use our models to create a catalogue of IMRI mergers proceeding on a step-by-step basis as follows: 
\begin{enumerate}
\item we divide the IMBH mass range -- $M_\ibh = 10^2-5\times 10^5\Ms$ -- in 15 values evenly distributed in logarithmic values;
\item for each IMBH mass, we create a sample of 100 stellar BHs with masses evenly drawn in the $M_\bh = 3.5-80 \Ms$ range;
\item we associate to the IMBH and BH a spin randomly drawn between 0-1;
\item we use \cite{jimenez17} numerical relativity fitting formulae to calculate the IMRI merger remnant mass $M_r$, spin $S_r$, and recoil velocity $v_r$, following the procedure depicted in \cite{arca20}.
\end{enumerate}
For each $M_\ibh -m_\bh$ pair, we repeat steps 3 and 4 100 times and, for each of them, we compare the recoil velocity with the escape velocity from the cluster centre, which we set to $\sigma = 5$ or $50$ km/s.

The result of this procedure is described in Figure \ref{fig:survival}, which shows the percentage of models for which $v_r < \sigma$. In the case of 
a ''normal'' cluster ($\sigma = 5$ km/s), we find that an IMBH with mass $M_\ibh < 200 \Ms$ gets ejected whenever the merging companion has a mass $m_\bh \geq 10 \Ms$. Even for heavier IMBHs, e.g. $M_\ibh = 1000\Ms$, the retention probability rapidly drops below $50\%$  if the companion mass exceeds $M_\bh > 20\Ms$. The retention probability attains values $>80\%$ regardless the secondary BH mass only for quite heavier IMBHs ($M_\ibh \sim 10^4\Ms$). The picture changes slightly in the case of a denser cluster. Assuming $\sigma=50$ km/s, we find that a moderate IMBH ($M_\ibh \sim 200-300 \Ms$) have a $50\%$ retention probability if $m_\bh < 30\Ms$, while for $M_\ibh = 1000\Ms$ the BH mass threshold to ensure a $50\%$ retaining probability increases to $m_\bh = 70\Ms$.

Our analysis suggests that localizing an IMBH in the centre of a star cluster and measuring its mass would enable us to get insights on the IMBH evolution. For instance, in the case of $M_\ibh > 10^3\Ms$ would suggest that the IMBH build-up was unlikely dominated by mergers with stellar BH companions, rather the major contribution came from accreted matter from disrupted stars.
}

@@

\subsection{IMBHs spin evolution}
{\bf

In this section we investigate whether the effects of a merging events can be encoded in the remnant IMBH spin.

To support this idea, we create a sample of IMBH-BH mergers based on our previous experiments as follows. We select an IMBH with mass of either $M_\ibh = 10^2 - 10^3 - 10^4 - 10^5$ and spin $S_\ibh = 0.1 - 0.5 - 0.9$ and we merge it with a stellar BH with mass in the range $3 - 60 \Ms$ drawn randomly and spin drawn accordingly to a Gaussian distribution with mean $S_\bh = 0.7$ -- i.e. the average value inferred from LIGO O1+O2 observations -- and dispersion 0.1.

@@

We see that as long as the mass ratio remains below $10^{-2}$ -- which implies an $M_\ibh > 10^4 \Ms$ -- the IMBH spin is not affected by the merger. However, for IMRIs with a mass ratio > 0.01, the remnant IMBH spin can change significantly, with a spin variation $\sim \pm 0.1$ for mass ratios in the range $q = 0.01-0.1$. At mass ratio $q>0.1$, the remnant IMBH spin ($S_r \sim 0.8\pm 0.2$) tends to attain a value close to the average value assumed for the BH spin distribution.
The peculiar dependence between the remnant spin and the IMRI mass ratio has important consequences for the IMBH evolution. One the one hand, if the IMBH formed from the collapse of a gaseous cloud or a very massive star born out of stellar collisions and reach masses $\gtrsim 10^4\Ms$, measuring its spin in a IMRI merger would provide us with information on the processes that regulated the IMBH very formation. On the other hand,
if the IMBH formed via multiple mergers in a ''BH-rich'' stellar environment, the remnant spin can provide clues on the spin distribution of the underlying population of stellar BHs.

To further investigate this point, we explore the effects of a series of subsequent mergers on the IMBH final spin. We assume $M_\ibh = 10^3\Ms$, i.e. a mass such to ensure a retention probability of at least $50\%$ in an IMRI merger event involving a BH smaller than $40\Ms$ (see Figures  \ref{fig:survival} and \ref{fig:f11a}). We set the IMBH spin to either $S_\ibh = 0.1-0.5-0.9$. For each spin value, we model ten consecutive mergers with stellar BHs with masses $3-60\Ms$ and spin drawn from a Gaussian distribution with average $S_\bh = 0.5$ and dispersion 0.1. The resulting merger sequence represents an {\it IMBH merger tree}. For each value of the spin we thus create 100 merger trees and calculate the IMBH spin at each successive merger as shown in Figure \ref{fig:f11b}.
We note that all models tend to converge toward the mean value of the adopted spin distribution for stellar BHs, thus suggesting that it could be possible to link the IMBH spin with the overall spin distribution of stellar BHs inhabiting the star cluster. In fact, complementing a measure of the IMBH spin with the spin distribution inferred from stellar BH mergers observed as GW sources can help in unravelling the IMBH evolution, whether it was driven by mergers with compact objects or by stellar feeding.
}

@@

\subsection{Gravitational wave signal}
{\bf
Our simulations encompass a wide range of IMRI models, in fact touching the layer of ``ordinary'' BH binaries (mass ratio $q>0.1$) and 
scratching the limit of extreme-mass ratio inspirals ($q < 3 \times10^{-5}$). This setup implies that the GW emission connected with our modelled mergers can populate a wide range of frequencies, from milli- to deca-Hz. Figure \ref{fig:f13} show the characteristic amplitude \citep{kocsis12,ASLiK18,seoane18,robson19} as a function of the frequency for a sample of IMRI mergers with different IMBH mass, assuming they are located at redshift $z=0.1$ ($\sim 460$ Mpc) and an observation time of 4 yr. We compare the modelled signal with the sensitivity curve of several detectors, namely the laser interferometer space antenna \citep[LISA,][]{eLISA13,amaro17lisa}, the advanced laser interferometer antenna \citep[ALIA,][]{bender13}, the decihertz observatory \citep[DO,]{arca19}, the Deciherz Gravitational-Wave Observator \citep[DECIGO,]{Kawamura11}, LIGO \citep{abbott16}, and the Einstein Telescope \citep[ET]{punturo10}. We see that mergers involving IMBHs with masses $< 10^5\Ms$ are promising multiband sources that can be seen in one detector during the inspiral phase and in another during the merger. The number of regions in which a detector is sensitive traversed by an IMRI widens at increasing the IMBH mass, in fact in principle an IMBH with mass $<10^3\Ms$ can be seen across the whole milli-decaHz frequency range.

@@

\subsection{Eccentricity of IMRIs}
{\bf
One important parameter that could be inferred from GW observations is the eccentricity of the source. To better understand whether our IMRI models are expected to retain a significant eccentricity while transiting across different observational bands, we show in Figure \ref{fig:f10} the average value of the eccentricity calculated when IMRIs cross the frequency windows $10^{-3}-10^{-1}-1$ Hz for all simulation sets. It is apparent that even when emitting low-frequency GWs, our modelled IMRIs have already significantly circularized, although apparently lower-mass IMRIs tend to have slightly larger eccentricities. The fact that our IMRIs are substantially circular when becoming observable to GW detectors has implications on their detectability, as we show in the next section.
}

@@

{\bf Over the last decade a number of works pointed out the potential presence of IMBHs in Galactic globular clusters, although in most cases the results were inconclusive. The family of clusters orbiting closer than 8 kpc to us include several IMBH host candidates: 47Tuc \citep{kiziltan17}, NGC6266 \citep{abbate19}, NGC6128 and NGC288 \citep{sollima16}, NGC6388 and NGC2808 \citep{miocchi07,lanzoni07}.

However, whether this clusters actually host an IMBH is a still open question. For instance, the inferred value for the 47Tuc IMBH mass goes from $M_\ibh > 2,000\Ms$ \citep{kiziltan17} to substantially lower values $\ll 1,000\Ms$ \citep{abbate19b}, depending on the observational technique adopted. A similar discussion involves other clusters, like NGC6388 \citep{lanzoni13,lutzgendorf13}. Therefore, LISA might play a crucial role in unravelling the presence (or not) of IMBHs in the Galaxy, at least in the closest ones.

LISA can also enable us to probe IMBHs even in nearby galaxies. For instance, it has been recently suggested that the Large Magellanic Cloud (LMC) might be harbouring an IMBH with a mass in the range $4\times 10^3-10^4\Ms$ \citep{erkal19} and up to $10^7\Ms$ \citep{boyce17}, while some of the clusters residing in the LMC can host IMBHs with masses in the range $10^3-10^4\Ms$ \citep{gualandris07}. Assuming a distance to the LMC $D_L = 49.97 \pm 1.13$ kpc \citep{pietrzy13}, LISA might deliver observations of {\it Magellanic} IMRIs with an SNR$= 8-40$, with the lower(upper) value corresponding to an IMBH with mass $10^3\Ms$($10^4\Ms$) and a BH companion of $30\Ms$. 
}

\subsection{IMRIs merger rate}

The launch of the LISA detector and the start of operations of the next generation of GW observatories like ET will enable us to unveil IMRIs at cosmological distances, provided that their number is sufficiently large to guarantee detection within the mission lifetime.
